### 2024-04-02 parts/intro.tex
typos

\AP There have been recurrent attempts to generalize the notions of regularity
from languages to functions, leading to classes of various expressiveness, such
as\footnote{This list is far from exhaustive.} Mealy Machines ($\intro*\Mealy$)
\cite{MEAL55}, sequential functions ($\intro*\Sequential$) \cite{SCHU77},
rational functions ($\intro*\Rational$) \cite{EILE74}, regular functions
($\intro*\Regular$) \cite{ENMA02}, and polyregular functions ($\intro*\Poly$)
\cite{BOKL19}. For all of these models, an \emph{aperiodic} counterpart can be
defined (respectively, $\intro*\AMealy$, $\intro*\ASequential$,
$\intro*\ARational$, $\intro*\ARegular$, $\intro*\SF$), lifting the
correspondence between counter-free automata, star-free languages and aperiodic
monoids to the functional setting
\cite{FKT14,BOJA14,CADA15,DJR16,BDK18,BOKL19,DGK21}. The inclusions between
these classes of functions are all known to be strict, and we depicted in
\cref{computational-models:fig} the status of the \emph{membership problem}
associated to these strict inclusions (that is, given a function $f$, decide if
$f$ can be computed by a function of a proper subclass). Note that in
\cref{computational-models:fig}, if $\mathsf{B}$ is a proper subclass of
$\mathsf{C}$, then the \emph{aperiodic} variant of $\mathsf{B}$ is precisely
the intersection of $\mathsf{B}$ with the \emph{aperiodic} variant of the
larger class $\mathsf{C}$. As an example, to conclude that a function can be
computed by an \emph{aperiodic sequential function} ($\ASequential$) it is
enough to prove that it is sequential ($\Sequential$), and computed by an
aperiodic polyregular function ($\SF$).

@@

\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
as having outputs in $\Nat$, and this is why they were introduced respectively
as \kl{$\Nat$-polyregular functions} ($\intro*\NPoly$) and \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$) \cite{DOUE21,DOUE22}. From this
point of view, $\NPoly$ is a subclass of a larger and well-studied class of
functions called (non-commutative) \intro{rational series}
\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
\kl{$\Nat$-polyregular functions} correspond to a natural subclass of
\kl{$\Nat$-rational series} that was frequently rediscovered in the literature
\cite{SCHU62,KRRC13}. In \cite{CDTL23}, the authors introduced the class of
\kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$) that generalize
\kl{$\Nat$-polyregular functions} by allowing negative outputs. This remains a
subclass of \kl{rational series}, more precisely, of the class of
\kl{$\Rel$-rational series} ($\intro*\ZRat$). For this class of functions the
aperiodic variant $\intro*\ZSF$ has been shown to be decidable \cite{CDTL23}.
Beware that concurrent (non-equivalent) notions of aperiodicity exists in the
research area of rational series \cite{REUT80,DRGA19,CDTL23}, and that we only
refer to the one applying to rational series of polynomial growths introduced
in \cite{CDTL23}.

@@

\begin{figure}
    \centering
    \includestandalone[width=7cm]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions,
        arranged along two axes. The first one is the complexity
        of the output alphabet ($\Nat$, $\Rel$, $\Sigma$). The second
        one is the allowed computational power
        (star-free polyregular functions, polyregular functions, 
        rational series).
        Arrows denote strict inclusions,
        and effectiveness (both in terms of decidability and of effective
        representation) is represented by thick double arrows. Inclusions that are
        suspected to be effective are represented using a dashed arrow together with a
        question mark.
    }
    \label{previously-known-inclusions:fig}
\end{figure}

>>>>

\AP There have been recurrent attempts to generalize the notions of regularity
from languages to functions, leading to classes of various expressiveness, such
as\footnote{This list is far from exhaustive.} Mealy Machines ($\intro*\Mealy$)
\cite{MEAL55}, sequential functions ($\intro*\Sequential$) \cite{SCHU77},
rational functions ($\intro*\Rational$) \cite{EILE74}, \intro{regular functions}
($\intro*\Regular$) \cite{ENMA02}, and \intro{polyregular functions} ($\intro*\Poly$)
\cite{BOKL19}. For all of these models, an \emph{aperiodic} counterpart can be
defined (respectively, $\intro*\AMealy$, $\intro*\ASequential$,
$\intro*\ARational$, $\intro*\ARegular$, $\intro*\SF$), lifting the
correspondence between counter-free automata, star-free languages and aperiodic
monoids to the functional setting
\cite{FKT14,BOJA14,CADA15,DJR16,BDK18,BOKL19,DGK21}. The inclusions between
these classes of functions are all known to be strict, and we depicted in
\cref{computational-models:fig} the status of the \emph{membership problem}
associated to these strict inclusions (that is, given a function $f$, decide if
$f$ can be computed by a function of a proper subclass). Note that in
\cref{computational-models:fig}, if $\mathsf{B}$ is a proper subclass of
$\mathsf{C}$, then the \emph{aperiodic} variant of $\mathsf{B}$ is precisely
the intersection of $\mathsf{B}$ with the \emph{aperiodic} variant of the
larger class $\mathsf{C}$. As an example, to conclude that a function can be
computed by an \emph{aperiodic sequential function} ($\ASequential$) it is
enough to prove that it is sequential ($\Sequential$), and computed by an
\intro{aperiodic polyregular function} ($\SF$).

@@

\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
as having outputs in $\Nat$, and this is why they were introduced respectively
as \kl{$\Nat$-polyregular functions} ($\intro*\NPoly$) and \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$) \cite{DOUE21,DOUE22}. From this
point of view, $\NPoly$ is a subclass of a larger and well-studied class of
functions called (non-commutative) \intro{rational series}
\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
\kl{$\Nat$-polyregular functions} correspond to a natural subclass of
\intro{$\Nat$-rational series} ($\intro*\NRat$) that was frequently
rediscovered in the literature \cite{SCHU62,KRRC13}. In \cite{CDTL23}, the
authors introduced the class of \kl{$\Rel$-polyregular functions}
($\intro*\ZPoly$) that generalize \kl{$\Nat$-polyregular functions} by allowing
negative outputs. This remains a subclass of \kl{rational series}, more
precisely, of the class of \intro{$\Rel$-rational series} ($\intro*\ZRat$). For
this class of functions the aperiodic variant $\intro*\ZSF$ has been shown to
be decidable \cite{CDTL23}. Beware that several non-equivalent notions of
aperiodicity coexist for rational series \cite{REUT80,DRGA19,CDTL23}, and that
we only refer to the one applying to rational series of polynomial growths
introduced in \cite{CDTL23}.

@@

\begin{figure}
    \centering
    \includestandalone[width=7cm]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions,
        arranged along two axes. The first one is the complexity
        of the output alphabet ($\Rel$, $\Nat$, $\Sigma$). The second
        one is the allowed computational power
        (\kl{star-free polyregular functions}, \kl{polyregular functions}, 
        \kl{rational series}).
        Arrows denote strict inclusions,
        and effectiveness (both in terms of decidability and of effective
        representation) is represented by thick double arrows. Inclusions that are
        suspected to be effective are represented using a dashed arrow together with a
        question mark.
    }
    \label{previously-known-inclusions:fig}
\end{figure}

### 2024-04-02 parts/prelim.tex
typos

\AP In the rest of this paper, $\Rel$ (resp. $\Nat$) denotes the set of
integers (resp. non-negative integers). The capital letters $\Sigma,\Gamma$
denotes fixed alphabets, i.e. a finite set of letters, and $\Sigma^*, \Gamma^*$
(resp. $\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
$\Sigma, \Gamma$. The empty word is written $\varepsilon \in \Sigma^*$. When $w
\in \Sigma^*$ and $a \in \Sigma$, we let $\card{w} \in \Nat$ be the length of
$w$, and $\card[a]{w}$ be the number of occurrences of $a$ in $w$. We assume
that the reader is familiar with the basics of automata theory, in particular
the notions of monoid morphisms, idempotents in monoids, monadic second-order
($\MSO$) logic and first-order ($\FO$) logic over finite words (see e.g.
\cite{THOM97}). Let us recall that a monoid $M$ is \intro(monoid){aperiodic}
whenever for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} =
x^n$.

\AP We also assume the reader to be familiar with rational series and weighted
automata, and refer to \cite{BERE10} for a comprehensive introduction. Let us
recall that one of the equivalent definitions of a $K$-rational series (where
$K$ is a ring) is given by weighted automata, which are non-deterministic
finite automata with transitions labelled by elements of $K$, and whose output
on a given word is the sum over all accepting paths of the products of the
weights along the path.

>>>>

\AP In the rest of this paper, $\Rel$ (resp. $\Nat$) denotes the set of
integers (resp. non-negative integers). The capital letters $\Sigma,\Gamma$
denotes fixed alphabets, i.e. a finite set of letters, and $\Sigma^*, \Gamma^*$
(resp. $\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
$\Sigma, \Gamma$. The empty word is written $\varepsilon \in \Sigma^*$. When $w
\in \Sigma^*$ and $a \in \Sigma$, we let $\card{w} \in \Nat$ be the length of
$w$, and $\card[a]{w}$ be the number of occurrences of $a$ in $w$. We assume
that the reader is familiar with the basics of automata theory, in particular
the notions of monoid morphisms, idempotents in monoids, monadic second-order
($\MSO$) logic and first-order ($\FO$) logic over finite words (see e.g.
\cite{THOM97}). As aperiodicity will be a central notion of this paper, let us
recall that a monoid $M$ is \intro(monoid){aperiodic} whenever for all $x \in
M$, there exists $n \in \Nat$ such that $x^{n+1} = x^n$.

\AP We also assume the reader to be familiar with \kl{rational series} and
weighted automata, and refer to \cite{BERE10} for a comprehensive introduction.
Let us recall that one of the equivalent definitions of a $K$-rational series
(where $K$ is a ring) is given by the functions from $\Sigma^*$ to $K$ computed
by weighted automata, which are non-deterministic finite automata with
transitions labelled by elements of $K$, and whose output on a given word is
the sum over all accepting paths of the products of the weights along the path.
